Toronto's Fire Incidents and Contributing Factors: An Exploratory Data Analysis

### 2024-04-18 paper/paper.qmd
update data test script and fix grammar mistakes

---
title: "Toronto's Fire Incidents and Contributing Factors: An Explorative Analysis"
author: 
  - Jimmy Luc
thanks: "Code and data are available at: [https://github.com/rex009x/fire-incidents-toronto](https://github.com/rex009x/fire-incidents-toronto)."
date: today
date-format: long
abstract: "This paper uses publicly available data from [Toronto Open Data](https://open.toronto.ca/) to identify contributing factors to fire incidents and their resulting financial losses. Further analysis of Toronto Fire Services' response time, area of origin, and various fire prevention measures as considering factors in the estimation. The results reveal a positive correlation between estimated dollar loss and possible causes of the incident, namely, mechanical/electrical failures yielding the highest loss. These findings bring urgency to fire safety as they highlight the importance of proper education, maintenance, and safety inspections. This analysis encourages homeowners, businesses, and property managers to take initiative in order to ensure the well-being of others."
format:
  pdf:
    fig-pos: "H"
toc: true
number-sections: true
bibliography: references.bib
---

@@

The motivation behind this analysis stems from the importance of fire safety. Beyond the protection of assets, it's about keeping people safe. To ensure the safety of fellow Canadians, it is imperative to comprehend the underlying patterns and determinants of fire incident. Analyzing factors such as the fire's area of origin, and the presence, functionality, and configuration of fire alarms. These additional variables are essential in understanding the dynamics of fire incidents and formulating informed strategies for fire prevention and mitigation.

Despite the abundance of data available, there remains a gap in analyzing the factors that contributing to the estimated dollar loss per fire incident in the Toronto area such as sociological factors. To address this gap, I employ a data-driven approach, utilizing explorative data analysis techniques to reveal patterns and relationships within the dataset. My methodology involves exploring various factors such as property use, possible causes of the incident (e.g., mechanical/electrical failures), and the presence of functioning fire alarms. Additionally, I incorporate the number of responding personnel from the Toronto Fire Services (TFS) as a variable to assess its impact on the estimated dollar loss.

@@

The methodologies behind the dataset's measurement is meticulously organized to account for scalability. Throughout the dataset, there are classifications methods used in scenarios where many instances require categorization. This is a decision that made with the concept of scalability in mind as there are far too many descriptors in the real world to account such that a singular description sufficiently and accurately describes the events of each incident. For example, regarding the area of origin, the raw data directly from Toronto's Open Data contains 74 unique classifications for an area of origin. The authors of the dataset hold the responsibility to properly assign the classification of each incident. This may result in inaccurate classifications without a clearly defined outline of what differentiates one from another.

One difficulty with measuring data concerning fire incidents is the nature of fire; fire is destructive and has the capacity to reduce tangible data to ash. The dataset has many instances of undetermined data which is likely related to nature of the incident. Taking a look at the types of smoke alarms at the locations of these incidents, nearly a fifth of all incidents recorded from 2011 to 2022 in Toronto are undetermined. While there may be numerous justifications for such values, it is difficult to analyze the data without disregarding a large portion of data, potentially leading to inconsistencies in the analysis and measurement errors.

The dataset's concern for privacy is a bit alarming in regards to the provided latitude and longitude data points. The dataset contains information regarding the incident's nearest intersection which is generally comprised of two streets. While this alone does not breach the privacy of the victims of the fire incident, the latitude and longitude measurements bring you directly to the location of the incident. This may not be a concern for those two reside within high-rise apartments, but when considering detached homes with unique home addresses, this does raise some privacy concerns.

@@

Looking at @fig-3, the plot illustrates the average dollar loss in damages from fire incidents across various categorized areas of origin. Each bar representing a distinct area of origin, with the height indicating the average amount in damages. The categories with the highest average amount in damages all share the fact they are all highly populated areas or require the utmost fire safety precaution. While the leading area of origin is "other/undetermined", it is largely because of the dataset was unable to classify the incident to one particular area hence the high estimated loss due to damages. The largest estimated dollar loss recorded in the dataset which is estimated at $50 million occurred on May 7, 2019 at the York Memorial Collegiate Institute [@citeLargestCostFire]. This is one example of an incident that is categorized under "other/undetermined" which resulted in a high dollar loss. As there are far too many classifications for large buildings that may serve many purposes, it is likely the reason such area of origin is leading.

@@

@fig-4 shows the number of each type of smoke alarms found present during fire incidents in Toronto from 2011 to 2022.  Each bar representing a specific type or configuration of smoke alarm, while the height corresponds to the number of fire incidents associated with that particular fire alarm. We can see wireless smoke alarms to be the lowest number of smoke alarms found at the location of the incidents with the remaining smoke alarm types being within a reasonable range of each other. No alarm/undetermined is the most frequent which is likely related to the nature of the emergency as fire alarms may have been destroyed as a result.
```{r graph4, fig.cap="Number of each type of smoke alarm present during fire incidents in Toronto", fig.asp=0.9}
#| echo: false
#| warning: false
#| message: false
#| label: fig-4

@@

A weakness of this analysis is the integrity of the dataset after cleaning and processing which is outlined in @sec-appendix-a. This analysis could be improved in regards to the consideration of sociological perspectives concerning peoples feelings towards fire safety. I believe discussing the awareness and sentiments towards fire safety and the steps they have taken to ensure the well-being for themselves and those around them. Including some survey or statistical evidence through a sociological lens would greatly benefit the overarching message of fire safety. Additionally, the analysis of the Toronto Fire Services response time to emergency calls could have been analyzed to determine whether or not it has changed in recent years as Toronto continues to increase in population. In @fig-2 we see that the most densely populated locations of Toronto also have the most concentrated number of fire incidents and incorporating some data model to predict the estimated dollar loss in populated areas would inform those within these areas of the dangers of fire.

>>>>

---
title: "Toronto's Fire Incidents and Contributing Factors: An Exploratory Data Analysis"
author: 
  - Jimmy Luc
thanks: "Code and data are available at: [https://github.com/rex009x/fire-incidents-toronto](https://github.com/rex009x/fire-incidents-toronto)."
date: today
date-format: long
abstract: "This paper uses publicly available data from [Toronto Open Data](https://open.toronto.ca/) to identify contributing factors to fire incidents and their resulting financial losses. Further analysis of Toronto Fire Services' response time, area of origin, and various fire prevention measures as considering factors in the estimation. The results reveal a positive correlation between estimated dollar loss and possible causes of the incident, namely, mechanical/electrical failures yielding the highest loss. These findings bring urgency to fire safety as they highlight the importance of proper education, maintenance, and safety inspections. This analysis encourages homeowners, businesses, and property managers to take initiative in order to ensure the well-being of others."
format:
  pdf:
    fig-pos: "H"
toc: true
number-sections: true
bibliography: references.bib
---

@@

The motivation behind this analysis stems from the importance of fire safety. Beyond the protection of assets, it's about keeping people safe. To ensure the safety of fellow Canadians, it is imperative to comprehend the underlying patterns and determinants of fire incidents. Analyzing factors such as the fire's area of origin, and the presence, functionality, and configuration of fire alarms. These additional variables are essential in understanding the dynamics of fire incidents and formulating informed strategies for fire prevention and mitigation.

Despite the abundance of data available, there remains a gap in analyzing the factors that contribute to the estimated dollar loss per fire incident in the Toronto area such as sociological factors. To address this gap, I employ a data-driven approach, utilizing exploratory data analysis techniques to reveal patterns and relationships within the dataset. My methodology involves exploring various factors such as property use, possible causes of the incident (e.g., mechanical/electrical failures), and the presence of functioning fire alarms. Additionally, I incorporate the number of responding personnel from the Toronto Fire Services (TFS) as a variable to assess its impact on the estimated dollar loss.

@@

The methodologies behind the dataset's measurement is meticulously organized to account for scalability. Throughout the dataset, there are classifications methods used in scenarios where many instances require categorization. This is a decision that was made with the concept of scalability in mind as there are far too many descriptors in the real world to account such that a singular description sufficiently and accurately describes the events of each incident. For example, regarding the area of origin, the raw data directly from Toronto's Open Data contains 74 unique classifications for an area of origin. The authors of the dataset hold the responsibility to properly assign the classification of each incident. This may result in inaccurate classifications without a clearly defined outline of what differentiates one from another.

One difficulty with measuring data concerning fire incidents is the nature of fire; fire is destructive and has the capacity to reduce tangible data to ash. The dataset has many instances of undetermined data which is likely related to the nature of the incident. Taking a look at the types of smoke alarms at the locations of these incidents, nearly a fifth of all incidents recorded from 2011 to 2022 in Toronto are undetermined. While there may be numerous justifications for such values, it is difficult to analyze the data without disregarding a large portion of data, potentially leading to inconsistencies in the analysis and measurement errors.

The dataset's concern for privacy is a bit alarming in regards to the provided latitude and longitude data points. The dataset contains information regarding the incident's nearest intersection which is generally composed of two streets. While this alone does not breach the privacy of the victims of the fire incident, the latitude and longitude measurements bring you directly to the location of the incident. This may not be a concern for those who reside within high-rise apartments, but when considering detached homes with unique home addresses, this does raise some privacy concerns.

@@

Looking at @fig-3, the plot illustrates the average dollar loss in damages from fire incidents across various categorized areas of origin. Each bar represents a distinct area of origin, with the height indicating the average amount in damages. The categories with the highest average amount in damages all share the fact they are all highly populated areas or require the utmost fire safety precaution. While the leading area of origin is "other/undetermined", it is largely because the dataset was unable to classify the incident to one particular area hence the high estimated loss due to damages. The largest estimated dollar loss recorded in the dataset, which is estimated at $50 million occurred on May 7, 2019 at the York Memorial Collegiate Institute [@citeLargestCostFire]. This is one example of an incident that is categorized under "other/undetermined" which resulted in a high dollar loss. As there are far too many classifications for large buildings that may serve many purposes, it is likely the reason such area of origin is leading.

@@

@fig-4 shows the number of each type of smoke alarms found present during fire incidents in Toronto from 2011 to 2022.  Each bar represents a specific type or configuration of smoke alarm, while the height corresponds to the number of fire incidents associated with that particular fire alarm. We can see wireless smoke alarms to be the lowest number of smoke alarms found at the location of the incidents with the remaining smoke alarm types being within a reasonable range of each other. No alarm/undetermined is the most frequent which is likely related to the nature of the emergency as fire alarms may have been destroyed as a result.
```{r graph4, fig.cap="Number of each type of smoke alarm present during fire incidents in Toronto", fig.asp=0.9}
#| echo: false
#| warning: false
#| message: false
#| label: fig-4

@@

A weakness of this analysis is the integrity of the dataset after cleaning and processing which is outlined in @sec-appendix-a. This analysis could be improved in regards to the consideration of sociological perspectives concerning people's feelings towards fire safety. I believe discussing the awareness and sentiments towards fire safety and the steps they have taken to ensure the well-being for themselves and those around them. Including some survey or statistical evidence through a sociological lens would greatly benefit the overarching message of fire safety. Additionally, the analysis of the Toronto Fire Services response time to emergency calls could have been analyzed to determine whether or not it has changed in recent years as Toronto continues to increase in population. In @fig-2 we see that the most densely populated locations of Toronto also have the most concentrated number of fire incidents and incorporating some data model to predict the estimated dollar loss in populated areas would inform those within these areas of the dangers of fire.
